---
title: "Exploring The Failures of The New York Knicks"
author: "Daniel Liu"
date: "27/06/2023"
output: 
  pdf_document:
    fig_caption: yes
    extra_dependencies: ["float"]
always_allow_html: true
---

```{r setup, warning=FALSE, error=FALSE, message=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stopwords)
library(tidyverse)
library(tidytext)
library(textdata)
library(lubridate)
library(plotly)
library(cowplot)
library(polite)
library(rvest)
library(janitor)
library(data.table)
library(rstatix)
webshot::install_phantomjs()
options(dplyr.summarise.inform = FALSE)      # suppress any summary info from grouping
```

# Introduction

As an avid New York Knicks fan, I will gain further insight into why the New York Knicks have disappointed over the past decade in the NBA. The insights I uncover will help uncover key areas of improvement which will hopefully guide the Knicks to winning more games. I will approach this task through the [Six Steps of Data Analysis Process](https://www.geeksforgeeks.org/six-steps-of-data-analysis-process/).

# Questions of Interest

1. How has the rise in popularity of the 3 point shot impacted the Knicks?
2. What other key statistical areas of improvement are important?
3. Does the team salary play a role in the success of a team?

# Preparing data

I pulled data from a multitude of sources in conducting my analysis.

1. [NBA Database](https://www.kaggle.com/datasets/wyattowalsh/basketball) is a dataset containing statistics on all 30 NBA teams, 4800+ NBA players and 60,000+ NBA games throughout history.
2. [NBA API](https://github.com/swar/nba_api) is an API I used to obtain data on NBA team statistics as opposed to having to wrangle the Kaggle datasets. 
3. [NBA Salaries](https://www.kaggle.com/datasets/jarosawjaworski/current-nba-players-contracts-history) is a dataset containing NBA player salaries between 2010-2020. As this analysis is conducted up until 2022, I scraped data from [spotrac](https://www.spotrac.com/nba/cap/2022/) to obtain NBA team salaries between 2021-2022.  

\noindent\rule{16.6cm}{0.4pt}

### Extracting data using nba_api in Python:

```{python echo = F, eval = F}
import pandas as pd
import requests
import numpy as np
from nba_api.stats.endpoints import leaguegamefinder
from nba_api.stats.endpoints import teamyearbyyearstats
from nba_api.stats.endpoints import franchiseplayers
from nba_api.stats.endpoints import teamestimatedmetrics
from nba_api.stats.static import teams
from nba_api.stats.static import players
```

Obtaining advanced metrics for all teams 2013 on wards:  

```{python eval = F}
seasons = ['2013-14', '2014-15', '2015-16', '2016-17', '2017-18', '2018-19', '2019-20',
'2020-21', '2021-22', '2022-23']
team_metrics = []
for season in seasons:
    tmp = teamestimatedmetrics.TeamEstimatedMetrics(season = season).get_data_frames()[0]
    tmp["YEAR"] = season[:4]
    team_metrics.append(tmp)
team_metrics = pd.concat(team_metrics, ignore_index = True)
team_metrics.to_csv("team_metrics.csv", index = False)
```

Obtaining basic team statistics from 2013-2022 

```{python eval = F}
def get_team_stats(individual_team_id: int):
    all_df = teamyearbyyearstats.TeamYearByYearStats(team_id = individual_team_id).get_data_frames()[0]
    all_df = all_df.tail(10)
    all_df["YEAR"] = all_df["YEAR"].apply(lambda year: year[:4])
    return all_df
```

```{python eval = F}
team_data = []
for team_id in team_ids:
    team_info = get_team_stats(str(team_id))
    team_data.append(team_info)
all_team_df = pd.concat(team_data, ignore_index = True)
all_team_df.to_csv("all_team_stats.csv", index = True)
```

Obtaining New York Knicks team stats from 2013-2022   

```{python eval = F}
NYK_stats_df = teamyearbyyearstats.TeamYearByYearStats(team_id = 1610612752)
NYK_stats_df = NYK_stats_df.get_data_frames()[0].tail(10)
NYK_stats_df["YEAR"] = NYK_stats_df["YEAR"].apply(lambda year: year[:4])
NYK_stats_df.to_csv("nyk_team_stats.csv", index = True)
```

\noindent\rule{16.6cm}{0.4pt}

### Importing required files

```{r results='hide'}
game_df <- read_csv(here::here("nba_data", "game.csv"),
                    show_col_types = F)
salaries <- read_csv(here::here("nba_data", "nba_salaries.csv"),
                     show_col_types = F)
play_by_play_df <- read_csv(here::here("nba_data", "play_by_play.csv"),
                            show_col_types = F)
```

```{r}
all_team_stats <- read_csv(here::here("nba_data", "all_team_stats.csv"),
                           show_col_types = F)
nyk_team_stats <- read_csv(here::here("nba_data", "nyk_team_stats.csv"),
                           show_col_types = F)
team_metrics <- read_csv(here::here("nba_data", "team_metrics.csv"),
                         show_col_types = F)
```

# Process

In this section, I process and add to existing dataframes to make it suitable for this analysis by:

* Making the NBA API data frames a consistent format with all other data frames.
* Manipulating formatting of team salaries and performing web scraping to retrieve up to date data.
* Filtering data to be within the analysis period (2013-2022).
* Initialize useful data frames for the future.

\noindent\rule{16.6cm}{0.4pt}


Cleaning column names from NBA API csv's

```{r results='hide'}
all_team_stats <- clean_names(all_team_stats)
nyk_team_stats <- clean_names(nyk_team_stats)
team_metrics <- clean_names(team_metrics)
```

I noticed there were 31 NBA teams as opposed to 30 in the all_teams_stats data frame

```{r}
length(unique(all_team_stats$team_name))
```

This likely means a team changed their name throughout this period. To find this team, I must identify teams that hadn't played the whole ten seasons under one name

```{r}
all_team_stats %>% 
  group_by(team_name) %>% 
  count() %>% 
  filter(n != 10)
```

As the Charlotte Bobcats became the Hornets in 2013 (i.e. team name change), I need to change the Bobcats name to the Hornets.

```{r results='hide'}
all_team_stats <- all_team_stats %>% 
  mutate(team_name = recode(team_name, "Bobcats" = "Hornets"))

team_metrics <- team_metrics %>% 
  mutate(team_name = recode(team_name, "Charlotte Bobcats" = "Charlotte Hornets"))
```




I combine individual player salaries into 30 NBA team salaries to keep the analysis on teams, not individual players

```{r results='hide'}
salaries <- salaries %>% 
  select(team, salary, season) %>% 
  filter(season >= 2013) %>% 
  filter(!(team %in% c("Fenerbahce Ulker Fenerbahce Ulker", "Maccabi Haifa Maccabi Haifa",
                       "null Unknown", "Madrid Real Madrid", "Charlotte Bobcats",
                       "New Orleans Hornets"))) %>% 
  group_by(team, season) %>% 
  summarize(salary = sum(salary))
```

Next, I perform web-scraping to get 2021 and 2022 NBA salary data

```{r results='hide'}
scrape_salaries <- function(html_link, year){
  page <- read_html(html_link)
  table <- page %>% 
    html_table() %>% 
    data.frame() %>% 
    as_tibble() %>% 
    select(Team, Total.Cap) %>% 
    rename("salary" = "Total.Cap",
         "team" = "Team") %>% 
    mutate(across("salary", ~gsub("\\$", "", .))) %>% 
    mutate(across("salary", ~gsub("\\,", "", .))) %>% 
    mutate(season = year)
  
  table$salary <- as.numeric(table$salary)
  
  returned_df <- salaries %>% 
    bind_rows(table)
  
  return(returned_df)
  
}

salaries <- scrape_salaries("https://www.spotrac.com/nba/cap/2022/", 2022)
salaries <- scrape_salaries("https://www.spotrac.com/nba/cap/2021/", 2021)
```

As the analysis is conducted between 2013-2022, I filter for data from 29/10/2013 (start of 2013 NBA season) till present

```{r results='hide', message=FALSE}
game_df$game_date <- as.Date(game_df$game_date)
game_df$season_id <- as.character(game_df$season_id)
game_df <- game_df %>%
  filter(game_date > "2013-10-29") %>% 
  mutate(year = substring(season_id, 2))

play_by_play_df <- game_df %>% 
  select(game_id, game_date) %>% 
  right_join(play_by_play_df, on = "game_id") %>% 
  filter(game_date > "2013-10-29")
```

Preparing data frame on each NBA team and their wins, losses and win percentage between 2013-2022

```{r}
team_wins_df <- all_team_stats %>% 
  select(team_id, team_city, team_name, year, gp, wins, losses, win_pct) %>% 
  mutate(win_pct = win_pct * 100,
         team_city = paste(team_city, team_name, sep = " ")) %>%
  rename("full_name" = "team_city") %>% 
  arrange(year)

knitr::kable(head(team_wins_df, 6), booktabs = T,
             caption = "Dataframe of NBA teams with their wins and lossess (2013-2022)")
```

# Analyze:

In this section, I will demonstrate the shortcomings of the Knicks over the past decade and identify key areas of improvement going forth.

\noindent\rule{16.6cm}{0.4pt}

### A disappointing decade

When you ask any NBA fan to name the worst teams from the past decade, chances are the Knicks will be named. It's not hard to see why considering the Knicks rank third last in win percentage over the past decade, winning less than 40% of their games.
```{r fig.cap = "Average team win % (2013-2022)"}
team_wins_df %>% 
  mutate(ToHighlight = ifelse(team_name == "Knicks", "yes", "no")) %>% 
  group_by(team_name, ToHighlight) %>% 
  summarise(avg_win_pct = mean(win_pct)) %>% 
  ggplot(aes(x = reorder(team_name, avg_win_pct),
             y = avg_win_pct,
             fill = ToHighlight)) +
  geom_col() + 
  labs(title = "Win percentage of each NBA team",
       x = "team",
       y = "win percentage") + 
  coord_flip() + 
  scale_fill_manual(values = c("yes" = "#E94B3CFF", "no" = "#2D2926FF")) + 
  theme_classic() + 
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, size = 23))
```

Not only are the Knicks failing to win, but they're playing less than inspiring basketball. 80% of words containing sentiment used by commentators when reacting to Knicks games were negative words such as "miss", "bad" and "terrible".
```{r message=FALSE}
knitr::kable(play_by_play_df %>% 
  filter(player1_team_abbreviation == "NYK" | player2_team_abbreviation == "NYK" |
           player3_team_abbreviation == "NYK") %>% 
  select(game_id, homedescription) %>% 
  na.omit() %>% 
  unnest_tokens(
    output = word,
    input = homedescription,
    token = "words"
  ) %>% 
  anti_join(get_stopwords()) %>% 
  inner_join(get_sentiments("bing")) %>% 
  count(sentiment) %>% 
  mutate(proportion = n / sum(n)), booktabs = T,
  caption = "Knicks sentiment amongst commentators (2013-2022)")
```

Compare this to a league average of 75% between 2013-2022 and you find that the Knicks are exciting with their style of play
```{r message=FALSE}
knitr::kable(play_by_play_df %>% 
  select(game_id, homedescription) %>% 
  na.omit() %>% 
  unnest_tokens(
    output = word,
    input = homedescription,
    token = "words"
  ) %>% 
  anti_join(get_stopwords()) %>% 
  inner_join(get_sentiments("bing")) %>% 
  count(sentiment) %>% 
  mutate(proportion = n / sum(n)), booktabs = T,
  caption = "League wide sentiment amongst commentators (2013-2022)")
```

The average NBA fan could have already predicted these results and so the question that should be asked is what can be done now and in the future to bring the Knicks glory over the next decade?


### The key is the three

The tiles highlighted by the red box in the correlation matrix show the correlation of common NBA statistics with win percentage. On the surface, one might think that points would yield the highest correlation with winning seeing as the goal of basketball is to score more points than your opponents.


Interestingly, the 3 point shot has the greatest correlation with winning at 0.58 among traditional NBA statistics
```{r fig.cap = "Correlation matrix of common NBA statistics", message=FALSE, warning=FALSE}
all_team_stats %>% 
  select(win_pct, fg3_pct, reb, ast, stl, tov, blk, pts) %>% 
  cor_mat() %>% 
  gather(-rowname, key = cor_var, value = r) %>% 
  ggplot(aes(x = rowname, y = cor_var, fill = r)) + 
  geom_tile() + 
  labs(x = "variables", y = "variables") + 
  scale_fill_gradient(low = "light yellow", high = "dark green") +
  geom_text(aes(label = r)) + 
  geom_segment(aes(x = 0.5,xend = 8.5,y = 7.5,yend = 7.5),colour="#FF4500", size = 1.5) + 
  geom_segment(aes(x = 0.5,xend = 8.5,y = 8.5,yend = 8.5),colour="#FF4500", size = 1.5) +
  geom_segment(aes(x = 0.5,xend = 0.5,y = 7.5,yend = 8.5),colour="#FF4500", size = 1.5) +
  geom_segment(aes(x = 8.5,xend = 8.5,y = 7.5,yend = 8.5),colour="#FF4500", size = 1.5)
```


It's clear that a moderate positive correlation exists between win percentage and 3 point percentage, suggesting that the better a teams 3 point percentage, the more likely that team will achieve a greater win percentage.
```{r fig.cap = "Season 3 point percentage against win percentage for 30 NBA teams (2013-2022)", message=FALSE}
fg3_corr <- substr(as.character(cor(all_team_stats$win_pct, all_team_stats$fg3_pct)), start = 0, stop = 5)
all_team_stats %>% 
  ggplot(aes(x = fg3_pct, y = win_pct)) + 
  geom_point(color = "#F8766D") +
  labs(title = "Effect of 3 Point % on Winning",
       x = "Three Point Percentage",
       y = "Win Percentage") +
  geom_smooth(method = "lm", se = FALSE, color = "#00BFC4") +
  annotate(geom = "text", label = paste("r = ", fg3_corr), color = "black", x = 0.33, y = 0.82, size = 6) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, size = 23))
```

With the above demonstrating how crucial 3 point percentage is in modern basketball, how have the Knicks performed in this key statistic?


Aside from two significant anomalies, the Knicks have been a below average 3 point shooting team throughout the period 2013-2022.


Only 16 of the 30 NBA teams will make the "playoffs" to compete for the championship. When considering the importance of being at-least league average in terms of wins to make the playoffs and the significance of the 3 point percentage on winning, the Knicks have clearly not placed enough of an emphasis on efficient 3 point shooting 2013-2022.
```{r eval=F, echo=F}
all_team_stats$year <- as.character(all_team_stats$year)
nyk_team_stats$year <- as.character(nyk_team_stats$year)
```


```{r fig.cap="New York Knicks against league average 3 point percentage", message=FALSE}
all_team_stats %>% 
  group_by(year) %>% 
  summarise(league_3p_pct = mean(fg3_pct)) %>% 
  left_join(nyk_team_stats, by = "year") %>% 
  select(year, league_3p_pct, fg3_pct, win_pct) %>% 
  rename("knicks_3p_pct" = "fg3_pct", "knicks_win_pct" = "win_pct") %>% 
  pivot_longer(cols = c("league_3p_pct", "knicks_3p_pct"), names_to = "type", values_to = "three_pct") %>% 
  ggplot(aes(x = as.factor(year), y = three_pct, col = type, group = type)) +
  geom_line() + 
  geom_point() +
  theme_classic() + 
  labs(title = "Change in 3 point percentage over time",
       x = "Year",
       y = "3 Point Percentage") +
  theme(plot.title = element_text(hjust = 0.5, size = 19))
```

Not only have the Knicks shot poorly from three over the past decade, but they also attempt less threes compared to league average. 


The largest difference between the Knicks attempted threes against league average was in 2019. In 2019, the Knicks shot 600 less threes than league average and only won 31.8% of their games that season, indicating the importance of shooting more threes and with greater efficiency.

```{r fig.cap="New York Knicks against league average 3 point attempts", message=FALSE}
all_team_stats %>% 
  group_by(year) %>% 
  summarise(league_3s_attempted = mean(fg3a)) %>% 
  left_join(nyk_team_stats, by = "year") %>% 
  select(year, league_3s_attempted, fg3a) %>% 
  rename("knicks_threes_attempted" = "fg3a") %>% 
  pivot_longer(cols = c("league_3s_attempted", "knicks_threes_attempted"), names_to = "type", values_to = "threes_attempted") %>% 
  ggplot(aes(x = as.factor(year), y = threes_attempted, col = type, group = type)) +
  geom_line() + 
  geom_point() +
  theme_classic() + 
  labs(title = "Change in 3 point attempts over time",
       x = "Year",
       y = "3 Pointers attempted") +
  theme(plot.title = element_text(hjust = 0.5, size = 19))
```

One encouraging sign I observed from fig 4 is that over the past few seasons, the Knicks have been trending at or above league average in terms of three point efficiency, perhaps explaining an above 50 win percentage from 2020 onwards.

```{r}
tail(nyk_team_stats, 3) %>% 
  summarise(win_pct = sum(wins) / sum(gp) * 100)
```

I recognize that 3 pointers are only one part of the game, but with how integral it is to the modern NBA, the Knicks ought to place greater emphasis on shooting the three ball more efficiency. This might occur through free agency signings, drafting players or training our current players to improve our three ball efficiency.


### Prioritize defence or offense?

Net rating/efficiency is a formula that measures how efficient a team is on both offense and defense. Offensive net rating measures how many points a team allows per 100 possessions. Given the opponent the ball 100 times, how many points are they likely to score? The same principle applies with defensive net rating.  


When exploring advanced NBA statistics, a 0.96 correlation between winning percentage and net efficiency stands out. This should come as no surprise because the more you score and less your opponents score per 100 possessions, the better a chance you have at winning more games over the course of an 82 game NBA season. What is more interesting is whether defense or offensive is more important as we might like to know which is more of a priority when considering signing or drafting different players.

```{r results='hide'}
adv_cor_df <- team_metrics %>% 
  select(w_pct, e_pace, e_reb_pct, e_oreb_pct, e_off_rating, e_net_rating,
         e_def_rating, e_tm_tov_pct, e_ast_ratio) %>% 
  cor_mat() %>% 
  gather(-rowname, key = cor_var, value = r)
```

There is a moderate correlation of 0.62 between winning percentage and offensive efficiency rating and a moderate correlation of -0.52 between winning percentage and defensive efficiency rating. It's interesting to note that a teams offensive capabilities are slightly more linked with success compared to defense.

```{r fig.cap="Correlation matrix of advanced NBA statistics"}
adv_cor_df %>% 
  ggplot(aes(x = rowname, y = cor_var, fill = r)) + 
  geom_tile() + 
  labs(x = "variables", y = "variables") + 
  scale_fill_gradient2(low = "#fb6767", high = "#3CB043") +
  geom_text(aes(label = r)) + 
  geom_segment(aes(x = 0.5,xend = 9.5,y = 9.5,yend = 9.5),colour="#FF4500", size=1.5)+ 
  geom_segment(aes(x = 0.5,xend = 9.5,y = 8.5,yend = 8.5),colour="#FF4500", size=1.5)+
  geom_segment(aes(x = 0.5,xend = 0.5,y = 9.5,yend = 8.5),colour="#FF4500", size=1.5)+
  geom_segment(aes(x = 9.5,xend = 9.5,y = 9.5,yend = 8.5),colour="#FF4500", size=1.5)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


From the two plots below in fig 7 a few key observation can be made:

* In 8 out of 10 seasons, the Knicks placed below league average in both offensive and defensive efficiency. Considering the relatively high correlation between winning and efficiency statistics, this could help in explaining the Knicks poor performance between 2013-2022.
* In no singular season have the Knicks been an above average team in both offensive and defensive efficiency. There might be a trade off between offense and defense, perhaps explained by the difficulty in finding elite players who can excel in both facets of basketball.

```{r echo=F}
team_metrics$year <- as.character(team_metrics$year)

offset <- 15

eff_rank_metrics <- team_metrics %>% 
  filter(team_name == "New York Knicks") %>% 
  select(year, e_off_rating_rank, e_def_rating_rank) %>% 
  mutate(e_off_rating_rank = e_off_rating_rank - offset,
         e_def_rating_rank = e_def_rating_rank - offset)
```

```{r fig.cap = "Knicks offensive and defensive efficiency ranking compared to league average of 15"}
plot1 <- eff_rank_metrics %>% 
  ggplot(aes(x = year, y = e_off_rating_rank)) + 
  geom_col(aes(fill = e_off_rating_rank<0), position = "dodge", col = "transparent") + 
  scale_y_continuous(labels = function (x) x + offset) +
  theme_classic() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, size = 15)) +
  geom_segment(x = 0.55, xend = 10.45, y = 0, yend = 0) +
  labs(title = "Knicks offensive efficiency rank against league average",
       y = "offensive efficiency rank")

plot2 <- eff_rank_metrics %>% 
  ggplot(aes(x = year, y = e_def_rating_rank)) + 
  geom_col(aes(fill = e_def_rating_rank<0), position = "dodge", col = "transparent") + 
  scale_y_continuous(labels = function (x) x + offset) +
  theme_classic() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5, size = 15)) +
  geom_segment(x = 0.55, xend = 10.45, y = 0, yend = 0) +
  labs(title = "Knicks defensive efficiency rank against league average",
       y = "defensive efficiency rank")

plot_grid(plot1, plot2, nrow = 2)
```

On the point of trade-offs, I would advise the Knicks to not give up one for the other but try to maximize their efficiency in total. As difficult as that sounds, both play a significant role in improving win percentage.

* The knicks saw a reduction in defensive ranking between 2020-2021, holding offensive ranking constant leading to a reduction in win percentage
* The knicks saw a significant improvement in offensive ranking between 2021-2022, with a smaller drop in defensive ranking, leading to an improvement in win percentage.

The above two points demonstrate just how vital having a good ranking in both categories is to win percentage.

```{r message=FALSE}
nyk_team_stats$year <- as.character(nyk_team_stats$year)
knitr::kable(nyk_team_stats %>% 
  tail(3) %>% 
  left_join(eff_rank_metrics) %>% 
  select(year, win_pct, e_off_rating_rank, e_def_rating_rank) %>% 
  mutate(e_off_rating_rank = e_off_rating_rank + 15,
         e_def_rating_rank = e_def_rating_rank + 15), booktabs = T,
  caption = "Effects of changing efficiency ranking on win percentage")
```

### Has the Knicks payroll hindered winning?

A common consensus in sports is that the more you spend, the better your team performs.  


```{r echo = F}
salary_and_wins <- salaries %>% 
  left_join(team_wins_df, by = c("team" = "full_name", "season" = "year"))  

knicks_salary_and_wins <- salary_and_wins %>% 
  filter(team == "New York Knicks")
```

Despite its validity in practically every sport, there is no pattern to validate that claim for the 30 NBA teams over the 10 seasons played between 2013-2022. Furthermore, the red dots, each of which indicate a salary and win percentage for one of the New York Knicks 10 seasons between 2013-2022 demonstrates no visible relationship either.

```{r fig.cap = "Team salary vs win percentage (2013-2022)", message=FALSE}
salary_and_wins %>% 
  ggplot(aes(x = salary/1000000, y = win_pct)) + 
  geom_point() +
  geom_point(data = knicks_salary_and_wins, aes(x = salary/1000000, y = win_pct),
             color = "red", size = 5, pch = 4) + 
  geom_point(data = knicks_salary_and_wins, aes(x = salary/1000000, y = win_pct),
             color = "red") +
  labs(title = "Team salary against win percentage",
       x = "Team Salary (millions of $)",
       y = "Win Percentage") +
  theme_classic() + 
  theme(plot.title = element_text(hjust = 0.5, size = 23))
```

Rather than focusing on how total monetary spending directly relates to win percentage, we can look into how the Knicks team salary has compared to league average between 2013-2022 and its impact on winning.

```{r echo=FALSE}
league_salaries <- salaries %>% 
  group_by(season) %>% 
  summarise(league_avg_salary = (sum(salary)/30) / 1000000)

knicks_salaries <- salaries %>% 
  filter(team == "New York Knicks") %>% 
  group_by(season) %>% 
  summarise(knicks_salary = sum(salary)/ 1000000)
```

The data shows that the Knicks have slowly reduced their team salary relative to league average between 2013-2022. Honing in on 2019-2022, the Knicks were well below league average in terms of spending and this was their best period in terms of winning, winning 47.8% of games compared to just 33.1% from 2012-2018.

```{r fig.cap = "Change in team salary (2013-2022)", message=FALSE}
league_salaries %>% 
  left_join(knicks_salaries) %>% 
  reshape2::melt(id.var = "season", variable.name = "Team") %>% 
  ggplot(aes(x = as.factor(season), y = value)) + 
  geom_bar(aes(fill = Team), position = "dodge", color = "black", stat = "identity") +
  theme_classic() + 
  labs(title = "Knicks change in salary against league average",
       x = "Year",
       y = "Salary (in millions of dollars)") +
  scale_y_continuous(expand = expansion(mult = 0)) +
  theme(plot.title = element_text(hjust = 0.5, size = 19))
```

```{r}
knitr::kable(summarise(head(select(nyk_team_stats, year, win_pct), 6),
                       win_pct = mean(win_pct)), booktabs = T, caption = "Knicks win percentage (2012-2018)")
knitr::kable(summarise(tail(select(nyk_team_stats, year, win_pct), 4),
                       win_pct = mean(win_pct)), booktabs = T, caption = "Knicks win percentage (2019-2022)")
```

This salary analysis suggests that by keeping team salary below league average, the Knicks have tended to perform better. This could be for a multitude of reasons including:

* **The development of young players drafted by the Knicks on low salaries**. With the emergence of RJ Barrett, Immanuel Quickley and Mitchell Robinson as reliable players for the Knicks over the past few years, the Knicks have been able to keep players on rookie contracts. A lower salary, at least in the short term for the Knicks suggests good development of young talent who contribute to winning as there's less need to spend big on veteran players.

* **Better management of player contracts**. Since 2018, the Knicks have signed all-star calibre players like Julius Randle and Jalen Brunson on bargain contracts. The identification of under-valued talent has led the Knicks to allow for flexibility in trades and free agency, something the Knicks lacked between 2012-2017.

# Act

To conclude, I will come up with my top three recommendations that I believe will help the Knicks win in the future based on my analysis.

1. Invest available salary and draft picks in obtaining ["Three 'n D"](https://www.hoopsbeast.com/what-is-a-3-and-d-player/) players who shoot the three pointer efficiently and play excellent defense. This is a sought-after mold that will add to three ball and defensive efficiency.
2. Focus equally on defense and offense, whether it be through particular signing players or on training drills. The greatest reward is through stellar play in both aspects.
3. Don't overspend on veterans in free agency but sign nice complementary pieces to Julius Randle and Jalen Brunson, allowing for flexibility in mid-season trades.