---
title: "Untitled"
author: "Daniel Liu"
date: "2023-06-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(stopwords)
library(tidyverse)
library(tidytext)
library(textdata)
library(lubridate)
library(cowplot)
```

Reading in all data:
```{r}
# Gives me information about draft history.
draft_history_df <- read_csv(here::here("nba_data", "draft_history.csv"))

# Gives me information about the game with its attendance and game time.
game_info_df <- read_csv(here::here("nba_data", "game_info.csv"))

# Important: Tells me information about all the games played.
game_df <- read_csv(here::here("nba_data", "game.csv"))
glimpse(game_df)

play_by_play_df <- read_csv(here::here("nba_data", "play_by_play.csv"))
glimpse(play_by_play_df)

# VERY USEFUL:
# Especially person_id, school, country, height (compare knicks height vs rest of league height), team_abbrev, dleague (were they from d_league),  draft_round.
common_player_info_df <- read_csv(here::here("nba_data", "common_player_info.csv"))
glimpse(common_player_info_df)

# Might be useful for matching team id with name for visualizations. Only 30 rows so very good for that.
team_df <- read_csv(here::here("nba_data", "team.csv"))
glimpse(team_df)

# Nba player salaries: Pre-pro
salaries <- read_csv(here::here("nba_data", "nba_salaries.csv"))
salaries <- salaries %>% 
  select(name, team, salary, season) %>% 
  filter(season >= 2013) %>% 
  filter(!(team %in% c("Fenerbahce Ulker Fenerbahce Ulker", "Maccabi Haifa Maccabi Haifa", "null Unknown",
                       "Madrid Real Madrid", "Charlotte Bobcats", "New Orleans Hornets")))

# Individual player data:
player_info <- read_csv(here::here("nba_data", "all_seasons.csv"))
```



# Pre-processing data:

Step 1: To give every dataframe a year, month and date_time to be able to filter on it:

NOTE: IMPORTANT TO FILTER ONLY FOR DATA from 2013-10-29 onwards!!! Start of the 2013/2014 NBA season.

This means to show that the Knicks are poor, I'm not comparing against old old data, but rather,
comparing **within** the years (i.e. comparing seasons where the knicks did good vs bad) AND
comparing **between** other teams in the same period, especially those that did good.

VERY IMPORTANT: COMPARING **WITHIN** THE YEARS, LOOKING AT WHY HAVE THE KNICKS STARTED TO IMPROVE SIGNIFICANTLY SINCE 2020?


```{r}
# Adding year to game:
game_df$game_date <- as.Date(game_df$game_date)   # Make it a date object, don't care about time.
game_df$season_id <- as.character(game_df$season_id)
game_df <- game_df %>%
  filter(game_date > "2013-10-29") %>% 
  mutate(year = substring(season_id, 2))

# Adding year to game_info:
game_info_df$game_date <- as.Date(game_info_df$game_date)
game_info_df <- game_info_df %>% 
  na.omit() %>% 
  mutate(year = year(game_date)) %>% 
    filter(game_date > "2013-10-29")

nrow(game_info_df) == length(unique(game_info_df$game_id))    # Checking every game is unique. Yes since no. rows = no. unique game_id.

# Adding the date to play_by_play through left_join:
game_with_year <- game_df %>% 
  select(game_id, game_date)
play_by_play_df <- play_by_play_df %>% 
  left_join(game_with_year, on = "game_id") %>% 
  filter(game_date > "2013-10-29")

```

For the analysis, I require information on each of the 30 teams wins and losses, both home and away and in total. 
As such data was not provided, create it myself.
```{r}
glimpse(game_df)
# Home team wins:
home_team_wins <- game_df %>% 
  rename(team_id = team_id_home, team_abbreviation = team_abbreviation_home) %>% 
  group_by(team_id, team_abbreviation) %>% 
  summarise(home_wins = sum(wl_home == "W", na.rm = TRUE),
            home_losses = sum(wl_home == "L", na.rm = TRUE),
            home_win_proportion = home_wins / (home_losses + home_wins))

home_team_wins

# Away team wins:
away_team_wins <- game_df %>% 
  rename(team_id = team_id_away, team_abbreviation = team_abbreviation_away) %>% 
  group_by(team_id, team_abbreviation) %>% 
  summarise(away_wins = sum(wl_away == "W", na.rm = TRUE),
            away_losses = sum(wl_away == "L", na.rm = TRUE),
            away_win_proportion = away_wins / (away_losses + away_wins))

away_team_wins

# Combining two df's for each team and add variables (total_wins) and (total_losses)
team_wins_df <- home_team_wins %>% 
  inner_join(away_team_wins, by = c("team_id", "team_abbreviation")) %>% 
  mutate(total_wins = home_wins + away_wins,
         total_losses = home_losses + away_losses,
         total_win_percentage = total_wins / (total_wins + total_losses) * 100)

team_wins_df
```


It's not secret the Knicks have performed poorly over the past decade. As seen below, the Knicks are the 3rd worst team with a below 50 win percentage
between 2013 and 2022.
```{r}
team_wins_df %>% 
  mutate(ToHighlight = ifelse(team_abbreviation == "NYK", "yes", "no")) %>% 
  ggplot(aes(x = reorder(team_abbreviation, total_win_percentage),
             y = total_win_percentage,
             fill = ToHighlight)) +
  geom_col() + 
  labs(title = "Win percentage of each NBA team (2013-2022)",
       x = "Team",
       y = "Win Percentage") + 
  coord_flip() + 
  scale_fill_manual(values = c("yes" = "#E94B3CFF", "no" = "#2D2926FF")) + 
  theme_classic() + 
  theme(legend.position = "none")
```

FURTHERMORE:

# Analysis 1: Sentiment Analysis

## Introduction: Fun analysis illustrating why the Knicks were poor through sentiment analysis:

This is the New York Knicks announcer sentiment. As seen 0.802 of words which
contain sentiment are negative, words like "miss", "bad" which imply Knicks games were
perhaps hard to watch with their poor performances throughout this period.

* Note: Might be a good place to make a user-defined function.
```{r}
play_by_play_df %>% 
  filter(player1_team_abbreviation == "NYK" | player2_team_abbreviation == "NYK" | player3_team_abbreviation == "NYK") %>% 
  select(game_id, homedescription) %>% 
  na.omit() %>% 
  unnest_tokens(
    output = word,
    input = homedescription,
    token = "words"
  ) %>% 
  anti_join(get_stopwords()) %>% 
  inner_join(get_sentiments("bing")) %>% 
  count(sentiment) %>% 
  mutate(proportion = n / sum(n))
```

Compare this to the league average (i.e. across all 30 teams):
Only a 0.750 negative sentiment. Knicks clearly aren't inspiring with their basketball.
```{r}
play_by_play_df %>% 
  select(game_id, homedescription) %>% 
  na.omit() %>% 
  unnest_tokens(
    output = word,
    input = homedescription,
    token = "words"
  ) %>% 
  anti_join(get_stopwords()) %>% 
  inner_join(get_sentiments("bing")) %>% 
  count(sentiment) %>% 
  mutate(proportion = n / sum(n))
```

But we already knew this, but the question is *WHY*?
What can be done in the future to bring the Knicks back into their glory days and win the support of the fans again?



# Analysis 2: Comparison of the Knicks 3P% against the rest of the leagues:

## It's no secret that there has been an explosion of 3 Pointers in the past few decades, with the likes of Steph 
## Curry and revoluntising the game from beyond the arc. In this section, I explore the performance and impact
## that the three point shot has had on the New York Knicks since 2013.

Firstly: Preparatory step. For each of the 30 teams, have their home_game_stats, away_game_stats and all_game_stats in a dataframe.

Explain how each team now has all their games played in all_games_stats as a tibble containing details on the game in the all_game_stats column.
```{r}
# Obtaining every game that every team has played in:
glimpse(game_df)
glimpse(team_df)

# Obtains a dataframe containing each teams home_game_stats in a column:
team_df <- team_df %>% 
  select(id, full_name, abbreviation) %>% 
  left_join(game_df, by = c("id" = "team_id_home")) %>% 
  nest(home_game_stats = c(-id, -full_name, -abbreviation))

team_df

# Adds onto the dataframe each teams away_game_stats in a column:
team_df <- team_df %>% 
  left_join(game_df, by = c("id" = "team_id_away")) %>% 
  nest(away_game_stats = c(-id, -full_name, -abbreviation, -home_game_stats))

# Adds onto the dataframe each teams away_game_stats in a column: For each team, combine their home_game_stats and away_game_stats
add_all_stats <- function(home_game_stats, away_game_stats){
  list(bind_rows(team_df$home_game_stats[1], team_df$away_game_stats[1]))
}

team_df

# Adds onto the dataframe a column which contains statistics on each game played by each team:\
team_df$all_game_stats <- NA    # Initialize empty column

for (x in 1:30) {
  team_df$all_game_stats[x] <- list(bind_rows(team_df$home_game_stats[x], team_df$away_game_stats[x]))
}

team_df
```



Secondly, for the New York Knicks (filtering), consider their games played between 2013-2022
```{r}
# A dataframe on all the New York Knicks Games Played between 2013-2022:
knicks_games <- team_df %>% 
  filter(abbreviation == "NYK") %>% 
  select(all_game_stats) %>% 
  unnest(cols = c(all_game_stats))
```

Look at the trend of 3 point % over time for the knicks against how many wins we had that season.
NOTE: Knicks abbreviation is "NYK"
```{r}
# Focus is on 3 points attempted + made:
knicks_threes <- knicks_games %>% 
  select(year, team_abbreviation_home, wl_home, fg3a_home, fg3m_home, team_abbreviation_away, wl_away, fg3a_away, fg3m_away, wl_home, wl_away)

knicks_threes

# Obtain a column containing the number of 3 pointers made + attempted per game for every Knicks game between 2013-2022:
knicks_threes <- knicks_threes %>% 
  mutate(
    knicks_3a = case_when(
    team_abbreviation_home == "NYK" ~ fg3a_home,
    TRUE ~ fg3a_away
  ),
    knicks_3m = case_when(
    team_abbreviation_home == "NYK" ~ fg3m_home,
    TRUE ~ fg3m_away
  ))

# Obtain a column containing whether the knicks won that game or not (1 = win, 0 = loss):
knicks_threes <- knicks_threes %>% 
  mutate(
    knicks_win = case_when(
    (team_abbreviation_home == "NYK") & (wl_home == "W") ~ 1,
    (team_abbreviation_home == "NYK") & (wl_home == "L") ~ 0,
    (team_abbreviation_away == "NYK") & (wl_away == "W") ~ 1,
    TRUE ~ 0
    ))


# Calculating for each year, the knicks 3P% by summing up 3's attempted / 3's made that year:
# Note: Missing 2017 Data. MAKE MENTION IN MY REPORT. DATA NOT SO GOOD?
knicks_threes <- knicks_threes %>% 
  group_by(year) %>% 
  summarise(knicks_3p_percent = (sum(knicks_3m) / sum(knicks_3a)) * 100,
            knicks_3p_attemped = sum(knicks_3a))
```

Compare the Knicks total 3 point % over 2013-2022 compared to the rest of the league.
```{r}
# Performing same steps as above but now on all teams to get a league average:
all_teams_threes <- game_df %>% 
  select(year, team_abbreviation_home, wl_home, fg3a_home, fg3m_home, team_abbreviation_away, wl_away, fg3a_away, fg3m_away)

all_teams_threes <- all_teams_threes %>% 
  mutate(total_game_fg3a = fg3a_home + fg3a_away,
         total_game_fg3m = fg3m_home + fg3m_away)

all_teams_threes

all_teams_threes <- all_teams_threes %>% 
  group_by(year) %>% 
  summarise(all_teams_3p_percent = (sum(total_game_fg3m) / sum(total_game_fg3a)) * 100,
            avg_teams_3p_attempted = sum(total_game_fg3a) / 30)

all_teams_threes
```

Creating a visualization of 3P% by year of league average vs knicks:

Plot: Knicks 3P% against league average: For most seaasons, apart from 3 outliers, much lower.
```{r}
knicks_threes %>%
  left_join(all_teams_threes) %>% 
  select(-knicks_3p_attemped, -avg_teams_3p_attempted) %>% 
  reshape2::melt(id.var = "year", variable.name = "Teams") %>% 
  ggplot(aes(x = as.factor(year),y = value, col = Teams, group = Teams)) + 
  geom_line() +
  geom_point() +
  theme_classic() + 
  labs(title = "Knicks 3P% Against League Average",
       x = "Year",
       y = "3 Point Percentage") + 
  ylim(33, 40)
```

```{r}
knicks_threes %>% 
  left_join(all_teams_threes) %>% 
  select(-knicks_3p_percent, -all_teams_3p_percent) %>% 
  reshape2::melt(id.var = "year", variable.name = "Teams") %>% 
  ggplot(aes(x = as.factor(year),y = value, col = Teams, group = Teams)) + 
  geom_line() +
  geom_point() +
  theme_classic() + 
  labs(title = "Knicks 3P Attempted Against League Average",
       x = "Year",
       y = "3 Pointers Attempted")
```


**CONCLUSION**: As seen in plot2, there has been a steady trend upwards in the number of 3's shot by NBA teams. 
However, the New York Knicks have been behind the ball in terms of 3 Pointers attempted compared to the league average.
Especially prevalant in 2018 where the Knicks 3P attempted plumetted way below league average. In this season,
the Knicks finished the season with an atrocious 17 wins, finishing last in the NBA that season. Perhaps the sudden increase in wins
could be explained by the significantly increased volume of threes shot under new head coach Tom Thibodeau, who joined the Knicks in 2020,
and led the Knicks to 2 playoffs births out of his 3 seasons in charge, even advancing to the conference semi-finals in the 2022 season.

As seen in plot1, the Knicks have been shooting poorly in terms of 3P% compared to the rest of the league, apart from the outliers in 2020 and
2013. This could be indicative of the Knicks poor performance but given the Knicks sudden rise to success from 2020 onwards, and the subsequent
decrease in 3P% by the Knicks, and in plot2, the significant rise in 3PA, it can be said that the Knicks are shooting more 3's, at a worse rate,
but are better off doing so.

I recognise that 3 Pointers are only one part of the game but with how integral it is to the modern NBA, it seems as if shooting more threes
one recipe that might contribute to success.


# Analysis 3: Checking salary of the Knicks players:
```{r}
league_salaries <- salaries %>% 
  group_by(season) %>% 
  summarise(league_salary = (sum(salary)/30) / 1000000)

knicks_salaries <- salaries %>% 
  filter(team == "New York Knicks") %>% 
  group_by(season) %>% 
  summarise(knicks_salary = sum(salary)/ 1000000)

league_salaries %>% 
  left_join(knicks_salaries) %>% 
  reshape2::melt(id.var = "season", variable.name = "Team") %>% 
  ggplot(aes(x = as.factor(season), y = value, col = Team, group = Team)) + 
  geom_line() +
  geom_point() +
  theme_classic() + 
  labs(title = "Knicks Total Salary Against League Average Total Salary",
       x = "Year",
       y = "Salary (in millions of dollars)")
```

**CONCLUSION**: Clearly the Knicks keep with the league trend in spending. 
However, ever since the 2017 CBA (Collective Bargaining Agreement), it seems the league salary has overtaken the Knicks salary.
Perhaps this is due to the 2017 CBA introducing higher max salaries. As the Knicks haven't had a player under a max contract
since Carmelo Anthony, who left in 2017, this could explain the Knicks trending under league average salary.
However, it can be seen that salary and the amount spent on players likely isn't a significant factor for the Knicks in terms of winning.
The Knicks have only performed well from 2020 onwards in this period, the only times the Knicks salaries were below league average.
This is likely because of the better management of players and player contracts.

Therefore, it can be said that better management of player contracts, allowing for flexibility in trades and free agency is key to improving the Knicks winning.

## What colleges are the best at finding good players and are the Knicks drafting from these colleges?

## Are Non-College (i.e. Other Team/International players) better than College players in terms of avg_net_rating?
* Compare the two and see if College or Non-College players make the bigger impact?
* Then see what demographic the knicks are primarily drafting (pie-chart). College or non-college. 
* If Non-college are making the bigger impact and Knicks are drafting more college than league avg, perhaps this is a recommendation.

## How good are the Knicks at finding second round gems (i.e. players with net ratings > 0: Make a positive impact on the floor)

Choice of statistic: Net Rating: Measures a team's point differnential per 100 persons (i.e. how many points does a player add/remove per 100 possessions)?
Basically, how much better off is a team with the player on the court compared to without?

```{r}
player_rating <- player_info %>% 
  select(player_name, season, net_rating) %>% 
  group_by(player_name) %>% 
  summarise(avg_net_rating = mean(net_rating))

draft_picks <- draft_history_df %>% 
  filter(season > 2012) %>% 
  select(person_id, player_name, season, round_number, round_pick, overall_pick, team_abbreviation, organization, organization_type) %>% 
  left_join(player_rating, by = "player_name")
draft_picks %>% 
  filter(team_abbreviation == "NYK")
  
  

```

## Home win percentage vs Away win percentage?
